\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-14 papers/jss/article.Rnw
intermediate commit cause i’m getting lunch
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@786 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built with
different programming languages used for the different phases of data
analysis -- collection, cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments and languages.  Each stage of such a data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.

Given these requirements, how do we safely and efficiently share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.

\emph{Comma-separated values} (CSV) files can be read and written by many
applications and so are often used for exporting tabular data.  However, CSV
files have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation and
parsing, possibly limited precision and ambiguities in the format involving
special characters.  \emph{JavaScript Object Notation} (JSON) is another
widely-supported format used mostly on the web that removes many of these
disadvantages, but it too suffers from being too slow to parse and also does
not provide strong typing between integers and floating point.  Because the
schema information is not kept separately, multiple JSON messages of the same
type needlessly duplicate the field names with each message.  Lastly,
\emph{Extensible Markup Language} (XML) is a well-established and widely-supported
protocol with the ability to define just about any arbitrarily complex
schema. However, it pays for this complexity with comparatively large and
verbose messages, and added complexities at the parsing side (which are
somewhat mitigated by the availability of mature libraries and
parsers).

A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
and BSON both have R interfaces, but % \citep{msgpackR,rmongodb}, but
% DE Why do we cite these packages, but not the numerous JSON packages?
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time, or when application logic and
requirement changes dictate update to the message format.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. Many sources compare data serialization formats
and show Protocol Buffers compare very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages

>>>>

Modern data collection and analysis pipelines involve collections
of components to enhance conrol of complex systems through
reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for various phases of data analysis -- collection, 
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
each environment or language. Every stage of such a data
analysis pipeline may produce intermediate results that need to be
stored or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?

Such systems require reliable and efficient exchange of intermediate
results between the individual components, using formats that are
independent of platform, language, operating system or architecture.
Most technical computing languages such as R, Julia, Java, and Python
include built-in support for serialization, but the default formats 
are usually language specific and thereby lock the user into a single
environment.

Traditionally, scientists and statisticians often use character seperated
text formats such as \texttt{CSV} \citep{shafranovich2005common} to 
export and import data. However, anyone who has ever used this will have
noticed that this method has many limitations: it is restricted to tabular 
datasets, lacks type-safety, and has limited precision for numeric values.
Moreover, ambiguities in the format itself frequently cause problems. 
For example the default characters used as seperator and decimal point 
are different in various parts of the world.
\emph{Extensible Markup Language} (\texttt{XML}) is another text-based 
well-established and widely-supported format with the ability to define
just about any arbitrarily complex schema \citep{nolan2013xml}. However,
it pays for this complexity with comparatively large and verbose messages,
and added complexities at the parsing side (which are somewhat mitigated
by the availability of mature libraries and parsers).
A more modern, widely used format is \emph{JavaScript Object Notation}
(\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}. This format is text-based as well and used mostly
on the web. Several R packages implement functions to parse and generate
\texttt{JSON} data from R objects. A number of \texttt{JSON} dialects has
been proposed, such as \texttt{BSON} and \texttt{MessagePack} which both
add binary support. However, these derivatives are not compatible with 
existing JSON software, and have not been widely adopted.

\subsection{Why Protocol Buffers}

- This paper introduces another format: protocol buffers
- unique combination of features that make it very suitable for numerical computing:
- binary, schema, versioned, mature, high quality cross language implementations
- we argue that (complex) statistical applications will benefit from using this format

%we should probably explain what a schema is%
Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. Many sources compare data serialization formats
and show Protocol Buffers compare very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages
